Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-08-30 ms.tex
Notes on discussion and summary section added
Changing \rfloor to \rvert in equation 33

\be
J_{\rm d} & = & 
\frac{(\sqrt{6}k^2L)/(16\pi^3 \Delta^3)}{ (kr/\Delta)^2 + (\sigma - \sigma_{\rm s})^2}.
\label{eq:Jd}
\ee

@@

A better approximation to the true solution has been derived by \citet{2006ApJ...649...14D}, who generalized the closed-form solution in slab geometry found in \citet{1990ApJ...350..216N}. They found a solution which satisfies a $J=0$ boundary condition at $r=R$. We generalize their solution to allow emission at frequency $\nu_{\rm s}$ away from line center. The result can be written as a sum over spatial modes,
\be
J_0 = \frac{\sqrt{6}}{16\pi} \frac{L}{R^2}\sum_{n=1}^{\infty}n\frac{\sin{\kappa_n r}}{\kappa_n r}\exp{\left(\frac{\kappa_n \Delta}{k}|\sigma - \sigma_s|\right)},
\ee
which evaluates to
\be
J_0 & = & \left( \frac{\sqrt{6}L}{32\pi^2 \Delta} \right)
\left( \frac{1}{Rr} \right)
\left( 
\frac{ \sin(\pi r/R) }{ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_{\rm s}) \right] - \cos(\pi r/R)}
\right)
\label{eq:J0}
\ee

@@

% TODO: Go through and tell them why you're shifting by integer multiples of tau0. Little abrupt here
% Motivate why we're doing the plot the way we are in the first few lines. Build it up for a first time reader. Add more motivation!
% Introduce this by saying the previous figures show that as we go to larger optical depth, the solution appears to get more and more accurate. To more quantitatively explore this, fig 5 shows...
% Write a thesis statement at the start of your paragraph and tell the reader what this is all about. It's good if the first line describes what the paragraph is about.
Fig. \ref{fig:xinit} shows how the boundary term scales with both $\tau_0$ and $\sigma_s$. The photon initialization frequency (in units of $\sigma$, as in the differential equation) is shifted in integer multiples of $\tau_0$. For clarity, only the positive side of the spectrum is shown. From Eq. \ref{eq:hbc_scaling}, the size of the correction should become smaller with larger optical depths, following a $(a\tau_0)^{-1/3}$ scaling. This factor has been scaled out of the figure such that solutions for different optical depths and the same source frequency $\sigma_s$ should show close agreement in scale on the figure if the relation holds. Indeed, the scaled solutions converge as the optical depth becomes larger, indicating general agreement with the $(a\tau_0)^{1/3}$ scaling. The remaining discrepancy present is due to the width of the Doppler core with respect to the peak of the spectrum; at lower depths, this causes the assumptions about the functional form of the line profile to break down. From this, we conclude that the errors introduced by the incorrect separation of variables in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others are proportional to $(a\tau_0)^{1/3}$, where $a$ is the ratio of natural to Doppler broadening and $\tau_0$ is the optical depth at line center. An additional amount of error of order $(a\tau_0)^{1/3}$ is introduced as the photon frequency $\sigma_s$ from line center is increased by a factor of $\tau_0$.

@@

Let us define the damping frequency to be $\gamma \equiv i\omega$, which is real-valued and positive for damped solutions. There are certain values of $\gamma$ for which a resonant response is produced which we will call $\gamma_{nm}$. Here, $m$ is an index that represents individual frequency eigenmodes, as opposed to $n$ which represents the spatial mode number. Near these frequencies, the solution will have the form 
\be
J(n,\sigma,-i\gamma) & \simeq \frac{ J_{nm}(\sigma) }{\gamma_{nm} - \gamma} + {\rm terms\ which\ vary\ slowly\ in\ \gamma}.
\ee
Obtaining the values of these eigenfrequencies and their corresponding eigenfunctions $J_{nm}(\sigma)$ allows us to obtain the contribution of this mode to the total solution via contour integration. 
\be
J(r,\sigma,t) & = &  -j_0(\kappa_n r)  \int \frac{d\gamma}{2\pi i} e^{-\gamma t}J(n,\sigma,-i\gamma)
\nonumber \\ & \simeq & j_0(\kappa_n r)  \int \frac{d\gamma}{2\pi i} e^{-\gamma t} \frac{ J_{nm}(\sigma) }{\gamma - \gamma_{nm}} 
\nonumber \\ & \simeq & j_0(\kappa_n r)  J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Here we have closed the integral along the imaginary $\gamma$ axis by a semi-circle at infinity, using the residue theorem to obtain the answer. The contribution from the semi-circle at infinity is zero. Summing over all spatial eigenmodes $n$ and over all temporal eigenmodes $m$ for a given $n$, we get the final form of the result
\be \label{eq:Jrsigmat}
J(r,\sigma,t) & = & \sum_{n=1}^\infty j_0(\kappa_n r)  \sum_m J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Beginning with Eq. \ref{eq:Jrsigmat}, we use the derivative
\be
\frac{dj_0(\kappa_n r)}{dr} \rfloor_R & =& \frac{d}{dr}  \rfloor_R \left( \frac{\sin(\kappa_n r)}{\kappa_n r} \right)
=  \left( \frac{\cos(\kappa_n R)}{R} - \frac{\sin(\kappa_n R)}{\kappa_n R^2} \right)\rfloor_R
\nonumber \\ & = & \frac{(-1)^n}{R},
\ee
to obtain the flux at the surface, which is
\be
F(R,\sigma,t) & =& - \frac{4\pi}{3k\phi} \frac{dJ(R,\sigma,t)}{dr} 
= - \frac{4\pi}{3k\phi R}  \sum_{nm} (-1)^n J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Multiplying by $4\pi R^2$ gives the energy per time per frequency
\be
\frac{dE}{dtd\nu} & = & - \frac{16\pi^2 R}{3k\phi}  \sum_{nm} (-1)^n J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\label{eq:dEdtdnu}
\ee
Integrating this over time yields a factor $-1/\gamma_{nm}$, and over $d\nu = \sqrt{3/2} \Delta^2 \phi d\sigma$, a ``sum rule" results
\be
1 & = &  \sqrt{ \frac{3}{2} } \frac{16\pi^2R\Delta^2}{3kE} \sum_{nm} (-1)^n \gamma_{nm}^{-1} \int d\sigma J_{nm}(\sigma).
\ee
This non-trivial expression provides a check on the values of $\gamma_{nm}$ and the functions $J_{nm}(\sigma)$. This expression can also be written as
\be
1 & =& \sum_{nm} P_{nm},
\label{eq:sumrule}
\ee
where the contribution of each eigenfunction is
\be \label{eq:pnmsoln}
P_{nm} & \equiv & \sqrt{ \frac{3}{2} } \frac{16\pi^2R\Delta^2}{3kE}  (-1)^n \gamma_{nm}^{-1} \int d\sigma J_{nm}(\sigma).
\ee
These coefficients can have either sign depending on the value of $n$. The size of each contribution scales roughly as $0.5/(m+1/8)^{2/3}$, with a very weak dependence on $n$. This indicates the need for a large number of $n$ and $m$ to converge, i.e., it takes roughly ten times as many $m$ modes to reduce the size of $P_{nm}$ by a factor of ${\sim}$5. Numerical convergence would be obtained only when the size of additional $P_{nm}$ are negligible in the sum.

@@

Integrating $dE/dt d\nu$ from Eq. \ref{eq:dEdtdnu} over all times should give a distribution for the emitted frequencies. It is expected that this time-integrated spectrum of the response to an impulse will agree well with the solution for the $J=0$ steady-state spectrum (Eq. \ref{eq:H0surf}). Dividing by the energy $E$, we find the probability distribution
\be \label{eq:spectrum}
P(\nu) & = &  \frac{16\pi^2 R}{3k\phi E}  \sum_{nm} (-1)^n \gamma_{nm}^{-1} J_{nm}(\sigma).
\ee
Integrating over $\nu$ gives unity according to the sum rule in Eq. \ref{eq:sumrule}. In Fig. \ref{fig:steadystate}, the resulting spectrum is shown for a sum up to $n=20$ and $m=500$ (labelled ``Time-integrated'') as compared with two analytic solutions: the steady-state $J=0$ solution with no time dependence (``Steady State'') and the result for summing \textit{only} spatial modes in the eigenfunction expansion with the time dependence ignored (``Partial Sum''). This graphic illustrates the physics modeled by each variable in the expansion. For example, the spatial modes contribute to the solutions' accuracy in the core of the line. If more spatial modes were calculated, the agreement between the lines extends further into the core. If additional frequency modes are calculated, faster-oscillating terms are incorporated into the sum over the imaginary frequency variable $\gamma$ which create more perfect cancellations with the lower-order terms, reducing the ``ringing'' seen in the time-integrated spectrum. Extending the calculation deep into the line core by adding additional spatial modes could have an impact on the accuracy of the escape time distribution, but this would only affect the distribution at late times since core photons are trapped for a longer duration before escaping. This was the motivation for choosing a comparatively low number of spatial eigenmodes with respect to the number of frequency eigenmodes calculated. As a result of this choice, we expect that our semianalytic approach will underestimate the number of escaping photon at late times as compared with the Monte Carlo.

@@

\subsection{Dependence on Source Frequency}

\section{Acceleration Schemes}

There are several methods that are commonly used to accelerate Monte Carlo radiation transfer calculations.  The simplest class of methods that is specific to resonance line transfer are core skipping methods \citep{1968ApJ...153..783A,2002ApJ...567..922A}.  These methods focus on Monte Carlo transfer in the wings of the line and avoid scatterings in the core. Line core photons are pushed to the wings by biasing the atomic velocity distribution to high velocities so the opacity is lower and escape is more rapid.

>>>>

\be
J_{\rm d} & = & 
\left(\frac{\sqrt{6}k^2L}{16\pi^3 \Delta^3}\right)\left(\frac{1}{(kr/\Delta)^2 + (\sigma - \sigma_{\rm s})^2}\right).
\label{eq:Jd}
\ee

@@

A better approximation to the true solution has been derived by \citet{2006ApJ...649...14D}, who generalized the closed-form solution in slab geometry found in \citet{1990ApJ...350..216N}. They found a solution which satisfies a $J=0$ boundary condition at $r=R$. We generalize their solution to allow emission at frequency $\nu_{\rm s}$ away from line center. The result can be written as a sum over spatial modes,
\be \label{eq:J0_sum}
J_0 = \frac{\sqrt{6}}{16\pi} \frac{L}{R^2}\sum_{n=1}^{\infty}n\frac{\sin{\kappa_n r}}{\kappa_n r}\exp{\left(\frac{\kappa_n \Delta}{k}|\sigma - \sigma_s|\right)},
\ee
which evaluates to
\be
J_0 & = & \left( \frac{\sqrt{6}L}{32\pi^2 \Delta} \right)
\left( \frac{1}{Rr} \right)
\left( 
\frac{ \sin(\pi r/R) }{ \cosh \left[ \frac{\pi \Delta}{k R} (\sigma - \sigma_{\rm s}) \right] - \cos(\pi r/R)}
\right)
\label{eq:J0}
\ee

@@

% TODO: Go through and tell them why you're shifting by integer multiples of tau0. Little abrupt here
% Motivate why we're doing the plot the way we are in the first few lines. Build it up for a first time reader. Add more motivation!
% Introduce this by saying the previous figures show that as we go to larger optical depth, the solution appears to get more and more accurate. To more quantitatively explore this, fig 5 shows...
% Write a thesis statement at the start of your paragraph and tell the reader what this is all about. It's good if the first line describes what the paragraph is about.
Fig. \ref{fig:xinit} shows how the boundary term scales with both $\tau_0$ and $\sigma_s$. The photon initialization frequency $\sigma_s$ is shifted in integer multiples of $\tau_0$. For clarity, only the $\sigma > 0$ side of the spectrum is shown. From Eq. \ref{eq:hbc_scaling}, the size of the correction should become smaller with larger optical depths, following a $(a\tau_0)^{-1/3}$ scaling. This factor has been scaled out of the figure such that solutions for different optical depths and the same source frequency $\sigma_s$ should show close agreement in scale on the figure if the relation holds. Indeed, the scaled solutions converge as the optical depth becomes larger, indicating general agreement with the $(a\tau_0)^{1/3}$ scaling. The remaining discrepancy present is due to the width of the Doppler core with respect to the peak of the spectrum; at lower depths, this causes the assumptions about the functional form of the line profile to break down. From this, we conclude that the errors introduced by the incorrect separation of variables in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others are proportional to $(a\tau_0)^{1/3}$, where $a$ is the ratio of natural to Doppler broadening and $\tau_0$ is the optical depth at line center. An additional amount of error of order $(a\tau_0)^{1/3}$ is introduced as the photon frequency $\sigma_s$ from line center is increased by a factor of $\tau_0$.

@@

Let us define the damping frequency to be $\gamma \equiv i\omega$, which is real-valued and positive for damped solutions. There are certain values of $\gamma$ for which a resonant response is produced which we will call $\gamma_{nm}$. Here, $m$ is an index that represents individual frequency eigenmodes, as opposed to $n$ which represents the spatial mode number. Near these frequencies, the solution will have the form 
\be
J(n,\sigma,-i\gamma) & \simeq \frac{ J_{nm}(\sigma) }{\gamma_{nm} - \gamma} + {\rm terms\ which\ vary\ slowly\ in\ \gamma}.
\ee
Obtaining the values of these eigenfrequencies and their corresponding eigenfunctions $J_{nm}(\sigma)$ allows us to obtain the contribution of this mode to the total solution via contour integration. 
\be
J(r,\sigma,t) & = &  -\sum_{n=1}^{\infty}j_0(\kappa_n r)  \int \frac{d\gamma}{2\pi i} e^{-\gamma t}J(n,\sigma,-i\gamma)
\nonumber \\ & \simeq & \sum_{n=1}^{\infty}j_0(\kappa_n r)  \int \frac{d\gamma}{2\pi i} e^{-\gamma t} \frac{ J_{nm}(\sigma) }{\gamma - \gamma_{nm}} 
\nonumber \\ & \simeq & \sum_{n=1}^{\infty}j_0(\kappa_n r)  J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Here we have closed the integral along the imaginary $\gamma$ axis by a semi-circle at infinity, using the residue theorem to obtain the answer. The contribution from the semi-circle at infinity is zero. Summing over all spatial eigenmodes $n$ and over all temporal eigenmodes $m$ for a given $n$, we get the final form of the result
\be \label{eq:Jrsigmat}
J(r,\sigma,t) & = & \sum_{n=1}^\infty j_0(\kappa_n r)  \sum_m J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Beginning with Eq. \ref{eq:Jrsigmat}, we use the derivative
\be
\frac{dj_0(\kappa_n r)}{dr} \bigg\rvert_R & =& \frac{d}{dr} \left( \frac{\sin(\kappa_n r)}{\kappa_n r} \right)\bigg\rvert_R
=  \left( \frac{\cos(\kappa_n R)}{R} - \frac{\sin(\kappa_n R)}{\kappa_n R^2} \right)
\nonumber \\ & = & \frac{(-1)^n}{R},
\ee
to obtain the flux at the surface, which is
\be
F(R,\sigma,t) & =& - \frac{4\pi}{3k\phi} \frac{dJ(R,\sigma,t)}{dr} 
= - \frac{4\pi}{3k\phi R}  \sum_{nm} (-1)^n J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Multiplying by $4\pi R^2$ gives the energy per time per frequency
\be
\frac{dE}{dtd\nu} & = & - \frac{16\pi^2 R}{3k\phi}  \sum_{nm} (-1)^n J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\label{eq:dEdtdnu}
\ee
Integrating this over time yields a factor $-1/\gamma_{nm}$, and over $d\nu = \sqrt{3/2} \Delta^2 \phi d\sigma$, a ``sum rule" results
\be
1 & = &  \sqrt{ \frac{3}{2} } \frac{16\pi^2R\Delta^2}{3kE} \sum_{nm} (-1)^n \gamma_{nm}^{-1} \int d\sigma J_{nm}(\sigma).
\ee
This non-trivial expression provides a check on the values of $\gamma_{nm}$ and the functions $J_{nm}(\sigma)$. This expression can also be written as
\be
1 & =& \sum_{nm} P_{nm},
\label{eq:sumrule}
\ee
where the contribution of each eigenfunction is
\be \label{eq:pnmsoln}
P_{nm} & \equiv & \sqrt{ \frac{3}{2} } \frac{16\pi^2R\Delta^2}{3kE}  (-1)^n \gamma_{nm}^{-1} \int d\sigma J_{nm}(\sigma).
\ee
These coefficients can have either sign depending on the value of $n$. The size of each contribution scales roughly as $0.5/(m+1/8)^{2/3}$, with a very weak dependence on $n$. This indicates the need for a large number of $n$ and $m$ to converge, i.e., it takes roughly ten times as many $m$ modes to reduce the size of $P_{nm}$ by a factor of ${\sim}$5. Numerical convergence would be obtained only when the size of additional $P_{nm}$ are negligible in the sum.

@@

Integrating $dE/dt d\nu$ from Eq. \ref{eq:dEdtdnu} over all times should give a distribution for the emitted frequencies. It is expected that this time-integrated spectrum of the response to an impulse will agree well with the solution for the $J=0$ steady-state spectrum (Eq. \ref{eq:H0surf}). Dividing by the energy $E$, we find the probability distribution
\be \label{eq:spectrum}
P(\nu) & = &  \frac{16\pi^2 R}{3k\phi E}  \sum_{nm} (-1)^n \gamma_{nm}^{-1} J_{nm}(\sigma).
\ee
Integrating over $\nu$ gives unity according to the sum rule in Eq. \ref{eq:sumrule}. In Fig. \ref{fig:steadystate}, the resulting spectrum is shown for a sum up to $n=20$ and $m=500$ (labelled ``Time-integrated'') as compared with two analytic solutions: the steady-state $J=0$ solution with no time dependence (``Steady State'') and the result for summing \textit{only} spatial modes in the eigenfunction expansion with the time dependence ignored (``Partial Sum''), as in Eq. \ref{eq:J0_sum}. This graphic illustrates the physics modeled by each variable in the expansion. For example, the spatial modes contribute to the solutions' accuracy in the core of the line. If more spatial modes were calculated, the agreement between the lines extends further into the core. If additional frequency modes are calculated, faster-oscillating terms are incorporated into the sum over the imaginary frequency variable $\gamma$ which create more perfect cancellations with the lower-order terms, reducing the ``ringing'' seen in the time-integrated spectrum. Extending the calculation deep into the line core by adding additional spatial modes could have an impact on the accuracy of the escape time distribution, but this would only affect the distribution at late times since core photons are trapped for a longer duration before escaping. This was the motivation for choosing a comparatively low number of spatial eigenmodes with respect to the number of frequency eigenmodes calculated. As a result of this choice, we expect that our semianalytic approach will underestimate the number of escaping photon at late times as compared with the Monte Carlo.

@@

\subsection{Emission Away From Line Center}

\section{Discussion}
% How can you apply the solution? Talk about it as an application of the problem that justifies going through the effort of implementing it. It reads like an introduction right now, but we can modify it to include less detail.
% Don't explain exactly what hybrid diffusion models are. Remove that paragraph. Put a ref in for Smith and Densmore - lot of applications to speed up lya, including core skipping & hybrid diffusion methods. Another method that's been widely applied is modified random walk methods, but htose require a solution of transfer in a sphere. So, the primary application of this calcuation is to develop that random walk model that will be used in future applicaitons. Keep reference to Ahn et al, since they did basically this but not in spherical symmetry.
% Discussion section: tlak about the differences between what we find and the other results from other papers. Add this as well. 
% First part of discussion - compare to previous work and summarize differences. Move some of the comparison with Dijkstra to the discussion.
% Second part - application of solution
% Move summary to discussion - currently written more like a discussion. 
% Write summary like it's a complement to the abstract. Cliffnotes version for people who don't want to read it.
% We re-examined previous solutions, noting that they use this incorrect boundary conditoin. We re-examined the problem with an appropriate bc, examine spectral and time dpeendence, key results are these, plan to use in a MC acceleration scheme in the future. Comment on comparison with MC. One, maybe two paragraphs.
There are several methods that are commonly used to accelerate Monte Carlo radiation transfer calculations.  The simplest class of methods that is specific to resonance line transfer are core skipping methods \citep{1968ApJ...153..783A,2002ApJ...567..922A}.  These methods focus on Monte Carlo transfer in the wings of the line and avoid scatterings in the core. Line core photons are pushed to the wings by biasing the atomic velocity distribution to high velocities so the opacity is lower and escape is more rapid.
